---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# SimBu

<!-- badges: start -->
[![tests](https://github.com/omnideconv/SimBu/actions/workflows/test.yml/badge.svg)](https://github.com/omnideconv/SimBu/actions/workflows/test.yml)
[![license](https://img.shields.io/badge/license-LGPLv3-blue.svg)](https://github.com/omnideconv/SimBu/blob/master/LICENSE.md)
[![docs](https://img.shields.io/badge/docs-pkgdown-blue.svg)](https://omnideconv.github.io/SimBu)
[![Codecov test coverage](https://codecov.io/gh/omnideconv/SimBu/branch/master/graph/badge.svg)](https://app.codecov.io/gh/omnideconv/SimBu?branch=master)
<!-- badges: end -->

The goal of SimBu is to simulate pseudo-bulk RNAseq datasets with variable cell-type fractions baed on public or private single-cell RNAseq datasets.

## Installation

To install the developmental version of the package, run:

```{r, eval=FALSE}
install.packages("devtools")
devtools::install_github("omnideconv/SimBu")
```

To install from Bioconductor:

```{r, eval=FALSE}
if (!require("BiocManager", quietly = TRUE)) {
  install.packages("BiocManager")
}

BiocManager::install("SimBu")
```

## Usage

Create a dataset-object with local data and simulate a pseudo-bulk dataset

```{r example, eval=FALSE}
library(SimBu)
# use local data to build dataset
dataset <- dataset(annotation = annotation_dataframe, count_matrix = expression_matrix, name = "test_dataset")
simulation <- simulate_bulk(data = dataset, scenario = "random", scaling_factor = "NONE")
```


For more detailed usage instructions, check out [Getting started](http://omnideconv.org/SimBu/articles/simulator_documentation.html). 

You can find more information on the simulation setup and mRNA bias in our [SimBu preprint](https://doi.org/10.1101/2022.05.06.490889). 
