---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# SimBu

<!-- badges: start -->
<!-- badges: end -->

The goal of SimBu is to simulate pseudo-bulk RNAseq datasets from public or private single-cell RNAseq datasets.

## Installation

To install this package, use the following command:

``` r
install.packages("devtools")
devtools::install_github("omnideconv/simulator") # auth_token=XXX for private repository
```

If you want to use the public data integration the package offers ([sfaira](https://theislab.github.io/Datasets) database), you additionally have to install sfaira into a new conda environment; follow these steps in your terminal:
(You need python 3.7+ to install sfaira!)

```
conda create --name sfaira python=3.9 pip
conda activate sfaira
/path/to/conda/envs/sfaira/bin/pip install git+https://github.com/theislab/sfaira.git@main #this ensures that the correct pip is used
```

## Usage

Create a dataset-object with local data and simulate a pseudo-bulk dataset

```{r example, eval=F}
library(SimBu)
# use local data to build dataset
dataset <- dataset(annotation = annotation_dataframe, count_matrix = expression_matrix, name = "test_dataset")
simulation <- simulate_bulk(data = dataset, scenario = "uniform", scaling_factor = "none")
```


